Estimation of number of runaway electrons per avalanche in Earth's atmosphere

### 2020-10-03 EPL-2020.03/main.tex
EPL second revision

\usepackage{textcomp}
\usepackage{multirow}
% \usepackage{trackchanges}

@@

\begin{document}
	\abstract{
		The connection between thunderstorms and relativistic runaway electron avalanches is an important topic that has attracted the attention of many researchers. Among other things, there are a lot of various simulations of the dynamics of electron avalanches. This article was written mostly in response to the article "The critical avalanche of runaway electrons" by Evgeny Oreshkin et al, which shows rather large numbers for an estimate of the number of runaway electrons, but it also contains the results of our own simulation and comparison with other papers.
	}

@@

	A recent paper \cite{Oreshkin_2018} presents the results of the 3D Monte-Carlo simulation of runaway avalanche development. The paper is focused on the avalanche-to-streamer transition. One of the main results of \cite{Oreshkin_2018} is the estimation of the number of electrons in one avalanche: $10^{17}$ - $10^{18}$. The value is significantly larger than values presented in other simulations~\cite{Gurevich:2001, dwyer2003fundamental,dwyer2011low}, which show about $10^6$ fast electrons and $10^{10}$ slow electrons per avalanche in normal conditions (air pressure, gas composition, etc.). The difference could be attributed to some non-standard approximations made in the article.

@@

	The simulation data was fitted by exponential function from Eq.~\ref{eq:exp} with $N_0 = 1$ and $l_a$ as a free parameter. The resulting value is $l_a \approx 85\un{m}$ without positrons, and $l_a \approx 78\un{m}$ with positrons. For comparison, Eq.~\ref{eq:dwyer}~(Dwyer's formula) predicts $l_a \approx 64\un{m}$ . Differences between our and Dwyer's result can be explained by the fact that Dwyer conducted simulation for normal condition and extrapolated results for another pressure. In addition, differences in the configuration of the Monte-Carlo model can contribute. Using received values for characteristic length, we can estimate the full number of runaway electrons in avalanches for different acceleration lengths. The results are presented in Table~\ref{tab:approx}. For reasonable acceleration cell lengths  $1200 - 1700\un{meters}$ it gives only $10^6-10^8$ runaway electrons, which is by many order less than $10^{17}$ - $10^{18}$ predicted by \cite{Oreshkin_2018}. $10^{17}$ - $10^{18}$~runaway electrons could be obtained only for cell lengths closer to $4000\un{meters}$ which are not possible in real atmospheric conditions~\cite[p.69-85]{rakov_uman}. It should be noted that we considered an avalanche created by a single seed electron, while cosmic rays are usually assumed to be a source of seed particles. But the flux of electrons of the required energies lies in the range of $10^3-10^4\un{m^{-2}s^{-1}}$~\cite{dwyer2003fundamental, sato2015analytical, sato2016analytical, dorman2013cosmic}. And taking into account that the movement of one avalanche through a cloud $1-1.5\un{km}$ length takes less than $10\un{\mu s}$ of the avalanche created by these electrons is quite separated in time and space. So that even if we take one single avalanche and a set of avalanches created with cosmic rays, then we still will not be able to achieve the value obtained by \cite{Oreshkin_2018}.

@@

	When discussing lighting breakdown, it is important to compute not only number of breakdown electrons, but also the total number of electrons and ions generated. This number could be very roughly estimated from energy conservation law.

	The number of ions produced by a single RREA can be estimated using the following logic. Firstly, all of the energy runaway electrons receive from the electric field is spent on ionization, on the production of new runaway electrons and on acceleration. Received energy in a segment $[z, z + dz]$ is approximately $N(z) e E d z$. Here $e$ is the electron electric charge, and all of the runaway electrons are considered to move parallel to the $z$ axis. Further, the energy, spent on runaway electron production is about $\frac{d N(z)}{d z} \varepsilon_{0} d z$, where $\varepsilon_{0}$ is the mean kinetic energy of a runaway electron for conditions under consideration. Substitution of $\varepsilon_0$ in this formula also takes into consideration the acceleration of runaway electrons. According to Babich \cite{Babich2001}, the mean electron energy of RREA can be estimated as $\varepsilon_0 = e(E - E_c) l_a$, where $E_c$ is the critical electric field value. Consequently, in the frame of Dwyer's e-folding length, the mean energy is simply derived as $\varepsilon_0 = 7.3\un{MeV}$~\cite{Dwyer2007}. Finally, the number of ions is calculated as the subtraction of received energy and energy wasted on runaway electron production, all divided by ionization potential. Ionization potential is considered to be $15\un{eV}$. Therefore, after integration the following estimation of the number of ions is derived:
	\begin{equation}
		N_{ions}(z) = \frac{e E l_a - \varepsilon_0}{I} \left( e^{\frac{z}{\l_a}} - 1\right).
	\end{equation}
	For $z = 1000\un{m}$ we get $N_{ions} \approx 10^7$.

@@

	Dwyer positron feedback mechanism \cite{Dwyer2012feedback} could, in theory, produce a significant increase in the number of runaway electrons. Dwayer's calculation independently verified in work~\cite{skeltved2014}, but the recent research shows that the feedback value in this mechanism was rather overestimated~\cite{npmdwyer}.

@@

	In this work, we calculated the number of runaway electrons in conditions expected in real thunderclouds (size of the strong field region up to $1200\un{m}$ and field strength up to $200\un{kV/m}$). The simulation itself covers only cell lengths up to 800~m, but it is clear that electron production could be extrapolated. The simulations provide an upper bound for the number of runaway electrons of about $10^6$ runaway electrons or $10^{10}$ total ionization per one seed electron for length about $1200\un{meter}$, which is in accordance with \cite{Gurevich:2001, Dwyer2013_radio} and contradicts to the value presented in \cite{Oreshkin_2018} (more than $10^{16}$ runaway electrons). Especially since Oreshkin’s report considers only electrons, and our results show that the introduction of gamma and positron generation increases the yield of runaway electrons.
	Still, even with those interactions, the resulting numbers are by several orders lower than the ones shown in \cite{Oreshkin_2018}. Although the origin of the discrepancy could be investigated further, it is clear that the non-relativistic particle dynamics model could not be used to describe the generation of enough ionization to produce a lightning discharge.

>>>>

\usepackage{textcomp}
\usepackage{multirow}

@@

\begin{document}
    \abstract{
        The connection between thunderstorms and relativistic runaway electron avalanches is an important topic that has attracted the attention of many researchers. Among other things, there are a lot of various simulations of the dynamics of electron avalanches. This article was written mostly in response to the article "The critical avalanche of runaway electrons" by Evgeny Oreshkin et al, which shows rather large numbers for an estimate of the number of runaway electrons. The results of our own simulation and comparison with other papers are presented.
    }

@@

    A recent paper \cite{Oreshkin_2018} presents the results of the 3D Monte-Carlo simulation of runaway avalanche development. The paper is focused on the avalanche-to-streamer transition \revision{and consideration of a critical avalanche (That is, an avalanche in which the self-field becomes comparable to the external field. This condition is similar to the criterion of the avalanche-to-streamer transition for a Townsend electron avalanche~\cite{Meek1940}. The self-field is determined not only by the number of electrons, but also by the transverse size of the avalanche, therefore the transition to the critical stage is determined also by the transverse diffusion of the runaway electrons~\cite{Dwyer2010}. But in our calculation, the self-field of the avalanche is not taken into account, because the avalanche does not reach the critical stage in our simulation.)}. One of the main results of \cite{Oreshkin_2018} is the estimation of the number of electrons in one avalanche: $10^{17}$ - $10^{18}$. The value is significantly larger than values presented in other simulations~\cite{Gurevich:2001, dwyer2003fundamental,dwyer2011low}, which show about $10^6$ fast electrons and $10^{10}$ slow electrons per avalanche in normal conditions (air pressure, gas composition, etc.). The difference could be attributed to some non-standard approximations made in the article.

@@

    The simulation data was fitted by exponential function from Eq.~\ref{eq:exp} with $N_0 = 1$ and $l_a$ as a free parameter. The resulting value is $l_a \approx 85\un{m}$ without positrons, and $l_a \approx 78\un{m}$ with positrons. For comparison, Eq.~\ref{eq:dwyer}~(Dwyer's formula) predicts $l_a \approx 64\un{m}$ . Differences between our and Dwyer's result can be explained by the fact that Dwyer conducted simulation for normal condition and extrapolated results for another pressure. In addition, differences in the configuration of the Monte-Carlo model can contribute. Using received values for characteristic length, we can estimate the full number of runaway electrons in avalanches for different acceleration lengths. The results are presented in Table~\ref{tab:approx}. For reasonable acceleration cell lengths  $1200 - 1700\un{meters}$ it gives only $10^6-10^8$ runaway electrons, which is by many order less than $10^{17}$ - $10^{18}$ predicted by \cite{Oreshkin_2018}. $10^{17}$ - $10^{18}$~runaway electrons could be obtained only for cell lengths closer to $4000\un{meters}$ which are not possible in real atmospheric conditions~\cite[p.69-85]{rakov_uman}. It should be noted that we considered an avalanche created by a single seed electron, while cosmic rays are usually assumed to be a source of seed particles. But the flux of electrons of the required energies lies in the range of $10^3-10^4\un{m^{-2}s^{-1}}$~\cite{dwyer2003fundamental, sato2015analytical, sato2016analytical, dorman2013cosmic}. And taking into account that the movement of one avalanche through a cloud $1-1.5\un{km}$ length takes less than $10\un{\mu s}$ of the avalanche created by these electrons is quite separated in time and space. So that even if we take one single avalanche and a set of avalanches created with cosmic rays, then we still will not be able to achieve the value obtained by \cite{Oreshkin_2018}. \revision{Of course, we are considering various sources (for example, an extensive air showers can give a huge number of secondary electrons, however, in the context of studying thunderstorms, it is rather an exotic event~\cite{murzin_2007_book, grieder_2010_book}), but the results of simulations in ~\cite{Oreshkin_2018} are given per one seed electron, and this value is therefore comparable.}

@@

    When discussing light\revision{n}ing breakdown, it is important to compute not only number of breakdown electrons, but also the total number of electrons and ions generated. This number could be very roughly estimated from energy conservation law.

    The number of ions produced by a single RREA can be estimated by a following way. Firstly, all of the energy runaway electrons receive from the electric field is spent on ionization, on the production of new runaway electrons and on acceleration. Received energy in a segment $[z, z + dz]$ is approximately $N(z) e E d z$. Here $e$ is the electron electric charge, and all of the runaway electrons are considered to move parallel to the $z$ axis. Further, the energy, spent on runaway electron production is about $\frac{d N(z)}{d z} \varepsilon_{0} d z$, where $\varepsilon_{0}$ is the mean kinetic energy of a runaway electron for conditions under consideration. Substitution of $\varepsilon_0$ in this formula also takes into consideration the acceleration of runaway electrons. According to Babich \cite{Babich2001}, the mean electron energy of RREA can be estimated as $\varepsilon_0 = e(E - E_c) l_a$, where $E_c$ is the critical electric field value. Consequently, in the frame of Dwyer's e-folding length, the mean energy is simply derived as $\varepsilon_0 = 7.3\un{MeV}$~\cite{Dwyer2007}. Finally, the number of ions is calculated as the subtraction of received energy and energy wasted on runaway electron production, all divided by ionization potential. Ionization potential is considered to be $15\un{eV}$. Therefore, after integration the following estimation of the number of ions is derived:
    \begin{equation}
    N_{ions}(z) = \frac{e E l_a - \varepsilon_0}{I} \left( e^{\frac{z}{\l_a}} - 1\right).
    \end{equation}
    For $z = 1000\un{m}$ we get $N_{ions} \approx 10^7$.

@@

    Dwyer positron feedback mechanism \cite{Dwyer2012feedback} could, in theory, produce a significant increase in the number of runaway electrons. Dwyer's calculation independently verified in work~\cite{skeltved2014}, but the recent research shows that the feedback value in this mechanism was rather overestimated~\cite{npmdwyer}.

@@

    In this work, we calculated the number of runaway electrons in conditions expected in real thunderclouds (size of the strong field region up to $1200\un{m}$ and field strength up to $200\un{kV/m}$). The simulation itself covers only cell lengths up to 800~m, but it is clear that electron production could be extrapolated. The simulations provide an upper bound for the number of runaway electrons of about $10^6$ runaway electrons or $10^{10}$ total ionization per one seed electron for length about $1200\un{meter}$, which is in accordance with \cite{Gurevich:2001, Dwyer2013_radio} and contradicts to the value presented in \cite{Oreshkin_2018} (more than $10^{16}$ runaway electrons per one seed particle). It is worth mentioning that Oreshkin’s report considers only electrons, meanwhile our results show that the introduction of gamma and positron generation increases the number of runaway electrons in one avalanche. 
    Still, even with those interactions, the resulting numbers are by several orders lower than the ones shown in \cite{Oreshkin_2018}. Although the origin of the discrepancy could be investigated further, it is clear that the non-relativistic particle dynamics model could not be used to describe the generation of enough ionization to produce a lightning discharge.
